---
title: "Vectors and types"
output: html
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## Datasets

Let's start by using R to create a dataset, which we will then save in our `data/` directory in a file called `feline-data.csv`. 
First, let's create the dataset in R using the `data.frame()` function:

```{r}
# define the cats data frame
cats <- data.frame(coat = c("calico", "black", "tabby"),
                    weight = c(2.1, 5.0, 3.2),
                    likes_string = c(1, 0, 1))
```

Saving `cats` as a .csv file:

```{r}
# save the cats data frame as a local csv file
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```

Loading the `cats` dataset from the csv file:

```{r}
# load the cats data frame from the csv file
cats <- read.csv(file = "data/feline-data.csv")
cats
```

## Working with `cats`

Let's extract some columns from `cats`:

```{r}
# extract the weight column using `$`
cats$weight
# extract the coat column using `$`
cats$coat
```

Each column in `cats` is a *vector*. 

There are lots of operations that we can do on vectors, such as:

```{r}
# add 2 to the weight vector
cats$weight + 2
```

```{r}
# multiply the weight and likes_string vectors together
cats$weight * cats$likes_string
```

But what about:

```{r}
#| error: true
# add the weight and coat vectors together:
cats$weight + cats$coat
```

What happened?

## Data Types


We can ask what type or "class" or "type" of data something is:

```{r}
# check the class() of the weight column
class(cats$weight)
```

Three main types:

- `numeric` (which encompasses `double` and `integer`), 

- `logical`, 

- `character` 


Let's identify the class of several values:

```{r}
# identify the class of: 3.14, TRUE, and "banana"
class(3.14)
class(TRUE)
class("banana")
```


Let's look at another version of the cats data with an extra row:

```{r}
# read in "data/feline-data_v2.csv"
cats_v2 <- read.csv(file="data/feline-data_v2.csv")
cats_v2
```

Let's check the class of the `weight` column

```{r}
# check the class of the weight column
class(cats_v2$weight)
```





The weight column is no longer numeric!

```{r}
#| error: true
# try to add 2 to the weight column
cats_v2$weight + 2
```

What happened?



## Vectors and Type Coercion




```{r}
# create a vector consisting of 1, 8, and 1.2 using c():
c(1, 8, 1.2)
```


The *columns of a data frame are also vectors*:

```{r}
# print out the weight column of the cats data frame
cats$weight
```





### Coercion by combining vectors

All entries in a vector must have the same type, so `c()` will *coerce* the type of each element to a common type. G

What do you think will be the type of the following vector?

```{r}
# create a vector containing 2, 6, and "3"
quiz_vector <- c(2, 6, "3")
```


What about

```{r}
# a vector containing "a" and TRUE
coercion_vector <- c('a', TRUE)
coercion_vector

# a vector containing 0 and TRUE
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```




### The type hierarchy


The coercion rules go: `logical` -> `numeric` -> `character`

```{r}
# create a vector with "a" and TRUE. What will its type be?
c("a", TRUE)
```


Forcing coercion using `as.numeric()`, `as.logical()`, etc:

```{r}
# create a vector with "0", "2", "4" (character values)
character_vector_example <- c('0', '2', '4')
character_vector_example
# coerce the vector to numeric using as.numeric()
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
# coerce the NUMERIC vector to logical using as.logical()
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```






In the cats data, coerce `likes_string` to logical (`TRUE`/`FALSE`)


```{r}
# print out likes_string column of cats data
cats$likes_string
# redefine the likes_string column as a logical vector
cats$likes_string <- as.logical(cats$likes_string)
# print out the modified cats data frame
cats
```




::: {.callout}
## Challenge 1

In this exercise, you will clean the cat data set from the chapter about type coercion.

Fill in the ____ gaps in the code below:

```{r}
#| eval: false
# Read data
cats <- read.csv("data/feline-data_v2.csv")
# 1. Print the data
_____

# 2. Show an overview of the table that prints out the type of each column
_____(cats)

# 3. The "weight" column has the incorrect data type __________.
#    The correct data type is: ____________.

# 4. Correct the 4th weight data point with the mean of the two given values
cats$weight[4] <- 2.35
#    print the data again to see the effect
cats

# 5. Convert the weight to the right data type
cats$weight <- ______________(cats$weight)

#    Calculate the mean to test yourself
mean(cats$weight)

# If you see the correct mean value (and not NA), you did the exercise
# correctly!
```


Solution:

```{r}
# Read data
cats <- read.csv("data/feline-data_v2.csv")
# 1. Print the data
cats

# 2. Show an overview of the table that prints out the type of each column
str(cats)

# 3. The "weight" column has the incorrect data type CHARACTER.
#    The correct data type is: NUMERIC.

# 4. Correct the 4th weight data point with the mean of the two given values
cats$weight[4] <- 2.35
#    print the data again to see the effect
cats

# 5. Convert the weight to the right data type
cats$weight <- as.numeric(cats$weight)

#    Calculate the mean to test yourself (you should get 3.16)
mean(cats$weight)

# If you see the correct mean value (and not NA), you did the exercise
# correctly!
```

:::

 
 
 
 
 
 
 







## Some basic functions for creating vectors

The combine function, `c()`, can also be used both to create a new vector as well as to append things to an existing vector:

```{r}
# create a new vector ab_vector with entries "a" and "b"
ab_vector <- c('a', 'b')
ab_vector
# add "z" to ab_vector
combine_example <- c(ab_vector, 'z')
combine_example
```








You can also make series of numbers using the `:` syntax as well as the `seq()` function:

```{r}
# create a vector with the numbers 1 to 10 using `:`
mySeries <- 1:10
mySeries
# print a vector with the numbers 1 to 10 using `seq()`
seq(10)
# print a vector with the numbers 1 to 10 using `seq()` with intervals of 0.1
seq(1, 10, by = 0.1)
```
 
 
 
 
### Head and tail
 
 
The `head()` and `tail()` functions show the first and last few entries of a vector, respectively.

```{r}
# create a vector of integers from 20 to 25
sequence_example <- 20:25
# look at the first 2 entries using head()
head(sequence_example, n = 2)
# look at the last 4 entries using tail()
tail(sequence_example, n = 4)
```



### Summarizing a vector

`length()` computes the number of entries in the vector:

```{r}
# compute the length of the vector above using `length()`
length(sequence_example)
```


`class()` reports the class/type of the values in the vector:

```{r}
# report the class of the vector above using `class()`
class(sequence_example)
```


If the vector is numeric, you can compute things like the sum and mean of the numeric entries:


```{r}
# compute the mean of the vector
mean(sequence_example) 
# compute the sum of the vector
sum(sequence_example)
```











## Factors

Categorical variables/columns with a finite set of options will often be formatted as factor. 

```{r}
# look at the coat categorical varaible from the cats data frame
cats$coat
```


The `coat` variable is currently formatted as a character variable

```{r}
# compute the class of the coat variable
class(cats$coat)
```


We can convert it to a factor using the `as.factor()` function:

```{r}
# convert the coat column from cats to a factor
cats$coat <- as.factor(cats$coat)
# check its class
class(cats$coat)
```



Let's take a look at the factor-formatted `coat` column:

```{r}
# take a look at the coat column
cats$coat
```


It looks very similar to the character format, but now our output tells us that there are "levels": "black", "calico", "tabby". 





### Converting factors to numeric: 

If we convert `coat` to a numeric type, it replaces each level with a number:

```{r}
# convert the coat column vector to numeric
as.numeric(cats$coat)
```

But if the factor levels are themselves numbers, the numeric information gets lost:

```{r}
# Create a factor version of the weight column
factor_weight <- as.factor(cats$weight)
factor_weight
# convert the factor weight column vector to a numeric vector
as.numeric(factor_weight)
```



