---
title: "Subsetting data frames"
output: html
embed-resources: true
editor_options: 
  chunk_output_type: console
---


# Data frames

Recall our `cats` data frame:

```{r}
cats <-  data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1, 5.0, 3.2),
                   likes_string = c(1, 0, 1))
cats
```


Each column of a data frame is a vector.







## Extracting columns from a data frame

### The dollar sign syntax `$`


```{r}
# Using `$`, extract the coat column

```

What type of object does this return?








### The square bracket syntax with one dimension `[j]`

We can extract individual columns using `[j]`

```{r}
# Using `[j]`, extract the first column 

```

What type of object does this return?





This `[j]` syntax also works with *named* indexing. 

```{r}
# Using `[j]`, extract the "coat" column by name

```








### The square bracket syntax with two dimensions `[i, j]`


```{r}
# Using `[i, j]`, extract the coat column (which is the first column)

```

What type of object does this return?






What does this syntax do?

`df[i, j]` extracts the entry in the `i`th row and the `j`th column

```{r}
# Using `[i, j]`, extract the entry in the third row and the first column

```





A blank `i` or `j` tells R to extract *all* of the rows or columns

- `df[, j]` will extract all rows for the jth column, and 

- `df[i, ]` will extract the ith row across all columns.


```{r}
# Using `[, j]` extract the second column

```

What type of object does this return?

```{r}
# Using `[i, ]` extract the second row

```

What type of object does this return? And why?








### The double bracket syntax with one dimension `[[]]`

One additional way to extract columns:

```{r}
# using [[]], extract the first column

```


The double bracket syntax also works with named indexing:

```{r}
# using [[]], extract the first column using its name: "coat"

```




Compare this with:

```{r}
cats[1]
cats["coat"]
```







::: {.callout}

## Challenge 1

There are several subtly different ways to extract observations and elements from data.frames:

- `cats[2]`
- `cats$weight`
- `cats["weight"]`
- `cats[1, 2]`
- `cats[, 2]`
- `cats[3, ]`

Try out these examples and explain what type of object is returned by each one.

:::






## Tip: Renaming data frame columns


Like vectors, data frames have column names, which can be accessed with the `names()` function.

```{r}
# using `names()` print out the names of the cats data frame 

```

You can also use `colnames()`

```{r}
# using `colnames()` print out the names of the cats data frame 

```





If you want to rename the second column of `cats`, you can assign a new name to the second element of `names(cats)`.

```{r}
# rename the second column of cats to "weight_kg"

# print out cats

```











## Extracting multiple columns (and rows)

Each of the square bracket syntax (but not the `$` syntax) options allow us to select multiple columns at once:

Can you guess what the following code will return?

```{r}
cats[1:2]
```


What about:

```{r}
cats[c(3, 1), 3]
```







## Using logical vectors to subset data frames

To subset the cats data frame to just the rows where the weight of the cats is at least 3:

```{r}
# compute a logical vector to identify which rows correspond to weight > 3

# use this logical vector to subset the cats data

```






Task: select only the rows where the *cat's coat is tabby or calico* and return just the values in the *"weight" column*:

```{r}
# subset to just the rows where the coat is "tabby" or "calico" and 
# print the values in the "weight" column

```










## Gapminder example


```{r}
# load the gapminder data

```

The gapminder dataset contains information on the population, life expectancy, GDP per capita for a range of country-year combinations:

```{r}
# look at the first 6 rows of the gapminder data

```

```{r}
# print the dimension of the gapminder data

```


::: {.callout}
## Challenge 2

Each of the following pieces of code are incorrect. Identify the error and fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957
  
```{r}
#| eval: false
gapminder[gapminder$year = 1957, ]
```

2. Extract all columns except 1 through to 4

```{r}
#| eval: false
gapminder[, -1:4]
```

3. Extract the rows where the life expectancy is longer the 80 years

```{r}
#| eval: false
gapminder[gapminder$lifeExp > 80]
```

4. Extract the first row, and the fourth and fifth columns (`continent` and `lifeExp`).

```{r}
#| eval: false
gapminder[1, 4, 5]
```

5. Advanced: extract rows that contain information for the years 2002 and 2007
  
```{r}
#| eval: false
gapminder[gapminder$year == 2002 | 2007, ]
```

:::



::: {.callout}

## Challenge 3
Suppose we want to pull out information from Southeast Asia, which includes the countries of Myanmar, Thailand, Cambodia, Vietnam, and Laos. 

Subset the data just to the rows that include the countries in Southeast Asia by:

1. Defining a logical vector that is `TRUE` for all of the countries in Southeast Asia and `FALSE` otherwise. 

2. Using this logical vector to extract just the rows for Southeast Asia from the gapminder dataset.


:::


::: {.callout}
## Challenge 4
1. Why does `gapminder[1:20]` return an error? How does it differ from `gapminder[1:20, ]`?

2. Create a new `data.frame` called `gapminder_subset` that only contains rows 1 through 9 and rows 19 through 23.

:::


