---
title: Lists
embed-resources: true
---



## Lists

Another data structure is the *list*.


List are very flexible because you can put anything you want in it: unlike a vector, the elements of a list can have different data types. For example:

```{r}
list_example <- list(1, "a", TRUE)
list_example
```


Like a vector, the "length" of a list corresponds to how many entries it contains:

```{r}
length(list_example)
```


When printing the object structure with `str()`, we see the data types of all
elements:

```{r}
str(list_example)
```

To retrieve one of the elements of a list, we use the **double bracket** notation:

```{r}
list_example[[2]]
```

The elements of lists also can have **names**, they can be given by prepending
them to the values, separated by an equals sign:

```{r}
another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE)
another_list
```

This results in a **named list**. Now we have a new function of our object!
We can access single elements by an additional way!

```{r}
another_list$title
```

as well as using named indexing in the double square bracket notation.

```{r}
another_list[["title"]]
```


Lists, it turns out, can become a lot more complicated than vectors. While each entry of a vector is just a single value, each entry of a list can be any type of object, including vectors and data frames. For example, the following list of length three contains three entries: a numeric vector, a data frame, and a single character value:



```{r}
# define the cats data frame:
cats <- data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1, 5.0, 3.2),
                   likes_string = c(1, 0, 1),
                   stringsAsFactors = FALSE)
```



```{r}
complicated_list <- list(vec = c(1, 2, 9),
                         dataframe = cats, 
                         single_value = "a")
complicated_list
```



::: {.callout}
## Challenge 6
Create a list of length two containing a (1) character vector containing the letters "x", "y", "z" and (2) a data frame with two columns that looks like this.

```{r}
#| echo: false
data.frame(name = c("Henry", "Hannah", "Harvey"), grade = c("A", "B", "C"))
```


Your list output should look like this:

```{r}
#| echo: false
list(c("x", "y", "z"),
     data.frame(name = c("Henry", "Hannah", "Harvey"), grade = c("A", "B", "C")))
```

::: {.callout collapse=true}
## Solution to Challenge 6

```{r}
list(c("x", "y", "z"),
     data.frame(name = c("Henry", "Hannah", "Harvey"), grade = c("A", "B", "C")))
```

:::
:::



### Data frames as a special case of a list


It turns out that a data frame is a special kind of a list. Specifically, *a data frame is a list of vectors of the same length*. 

This is why you can extract vector columns from a data frame using the double brackets notation:

```{r}
cats
```

```{r}
cats[["coat"]]
```


Note that the `df[i, j]` index notation is specific to data frames (and does not work for lists).




### List subsetting

Now we'll introduce some new subsetting operators. There are three functions
used to subset lists. We've already seen these when learning about atomic vectors and matrices:  `[`, `[[`, and `$`.

Using `[` will always return a list. If you want to *subset* a list, but not
*extract* an element, then you will likely use `[`.

```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(mtcars))
xlist[1]
```

This returns a *list with one element*.

We can subset elements of a list exactly the same way as atomic
vectors using `[`. Comparison operations however won't work as
they're not recursive, they will try to condition on the data structures
in each element of the list, not the individual elements within those
data structures.

```{r}
xlist[1:2]
```

To extract individual elements of a list, you need to use the double-square
bracket function: `[[`.

```{r}
xlist[[1]]
```

Notice that now the result is a vector, not a list.

You can't extract more than one element at once:

```{r, error=TRUE}
xlist[[1:2]]
```

Nor use it to skip elements:

```{r, error=TRUE}
xlist[[-1]]
```

But you can use names to both subset and extract elements:

```{r}
xlist[["a"]]
```

The `$` function is a shorthand way for extracting elements by name:

```{r}
xlist$data
```

::: {.callout}
## Challenge 5

Given the following list:

```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(mtcars))
xlist
```

Using list and vector subsetting, extract the second entry in the second element of the list (i.e., extract the number `2` from the `b` entry from xlist`). There are several ways to do this. Compare your answer with your neighbor. Did you do it the same way?

::: {.callout collapse=true}

## Solution to challenge 5

```{r}
xlist$b[2]
```

```{r}
xlist[[2]][2]
```

```{r}
xlist[["b"]][2]
```
:::
:::